The polynomial representation of the type $A_{n - 1}$ rational Cherednik algebra in characteristic $p \mid n$

### 2015-05-15 formalwriteup.tex
more section 3 edits

 \newtheorem*{thm:main}{Theorem \ref{thm:main}}
\begin{thm:main} The irreducible representation $A/J$  of $\HH_{\hbar,c}(\h)$ is a complete intersection with  Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$. The maximal ideal $J$ is generated by $[z^p]F_a(z)$ for the formal power series
\[
F_a(z)=\frac{1}{1-x_az} \prod_{j=0}^{n-1}\left( \sum_{k=0}^{p-1} \binom{c}{k}(-1+\prod_{j=0}^{n-1} (1-x_jz))^k\right) \text{ for $a=0,\dots,n-2$}.
\]
\end{thm:main}

@@

Therefore we have $[z^p]F_j(z)=f_j=\sum_{k=0}^{(p-1)/2} (-1)^k\binom{c}{k}=\binom{c-1}{(p-1)/2}$ and $[z^p]F_i(z)=f_i=0$ for $n-1>i \ne j$ (since $p > 2$). Since $c$ is generic, we can assume $\binom{c-1}{(p-1)/2} \ne 0$. Then since $\sum_{i=0}^{n-2} \lambda_if_i=0$, we see that $\lambda_j\binom{c-1}{(p-1)/2}=0$; then we must have $\lambda_j=0$.

@@

\begin{proposition}\label{prop:ci} Let $I \subseteq A$ be the ideal generated by $f_i$ for $i=0,\dots,n-2$.  The $A/I$ is a complete intersection. 
\end{proposition}

@@

We suppose that the $x_i$ take the distinct values $\{y_0,\dots,y_{r-1}\}$ with multiplicity $m_i>0$; then since $\sum_{i=0}^{n-1} x_i=0$ we have $\sum_{i=0}^{r-1} m_iy_i=0$.

Then $g(z)=(1-y_iz)^{m_i}$. Recall that
\[
F(z)=\sum_{k=0}^{p-1} \binom{c}{k}(g-1)^k.
\]

We assume that $y_i \ne 0$ for all $i$.

Define
\[
a(z)=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z).
\]
Recall that we showed
\[
F'(z)=V(z)+\left(\sum_j \frac{-cx_j}{1-x_jz}F(z)\right)=V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right),
\]
where $[z^\ell]V(z)=0$ for $0 \le \ell \le p$. We then see that
\begin{align*}
a'(z)&=\frac{d}{dz}\left(\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z)\right)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F'(z)+\sum_{j=0}^{r-1}\frac{y_j}{(1-y_jz)\prod_{i=0}^{r-1}(1-y_iz)}F(z)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}\left(V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}a(z)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}-\left(\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)\right).
\end{align*}
Since $[z^\ell]V(z)=0$ for $0 \le \ell \le p$ we see that $\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}=z^pb(z)$ for some power series $b(z)$. Therefore
\[
a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)
\]
for some power series $b(z)$.

@@

Note that for any $i$, $a(z)b_i(z)=\frac{1}{1-y_iz}F(z)$. Then in particular, we have 
\[
[z^p]a(z)b_i(z)=[z^p]\frac{1}{1-y_iz}F(z)=0
\]
Therefore for any $\lambda_i \in k$, we have
\begin{equation}\label{eq:lincomb}
\sum_{i=0}^{r-1} \lambda_i[z^p]a(z)b_i(z)=[z^p]\left( a(z)\sum_{i=0}^{r-1}\lambda_ib_i(z)\right)=0.
\end{equation}
Recall that the $b_i(z)$ are linearly independent; they also have degree at most $r-1$. Therefore for $0 \le k \le r-1$ we can choose $\{\lambda_i^k\}$ so that 
\[
\sum_i\lambda_i^kb_i(z)=z^k.
\]
Then if we use the $\lambda_i^k$ in Equation~\ref{eq:lincomb}, we see that $[z^p](a(z)z^k)=[z^{p-k}]a(z)=0$ for $k=0,\dots,r-1$.

We will show that we must have $[z^0]a(z)=0$, a contradiction. Note that if $r > p$ then we have already shown that; therefore we assume $r \le p$.

For any integer $\ell$ we define the Laurent polynomial 
\[
h_\ell(z)=z^{-\ell-r}\prod_i(1-y_iz).
\]
Then we have
\begin{align*}
\frac{d}{dz}( h_\ell(z) a(z)) &= -(l + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) + h_\ell(z) a'(z) \\
&=  -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) - h_\ell(z) \sum_j \frac{(m_j c - 1)y_i}{1 - y_ijz} a(z)  + z^ph_\ell(z) b(z)\\
&= -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_im_i c}{1 - y_iz} h_\ell(z) a(z) + z^ph_\ell(z) b(z)\\
&= -\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z) a(z) + z^ph_\ell(z) b(z).
\end{align*}
We note that 
\[
-\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z)
\]
is a Laurent polynomial with lowest degree term 
\[
- z^{-l-r-1} \left(\ell + r + \sum_i y_i m_i c\right) = - z^{-\ell-r-1}(\ell+ r)
\]
and highest degree term 
\[
- z^{-\ell - 1} (-1)^r  \prod_i y_i\left((\ell + r) - \sum_i m_i c\right).
\]
The lowest degree term of the formal power series $z^ph_\ell(z)b(z)$ has degree at least $p-\ell-r$. We note that we must have $[z^{-1}]\left(\frac{d}{dz}( h_\ell(z) a(z))\right)=0$.

Then as long as $p-\ell-r \ge 0$, which is true when $\ell \le p-r$, we see that $[z^{-1}]\left(\frac{d}{dz}( h_\ell(z) a(z))\right)=0$ is a linear combination of $[z^\ell]a(z),[z^{\ell+1}]a(z),\dots,[z^{\ell+r}]a(z)$ in which $[z^\ell]a(z)$ has nonzero coefficient $ (-1)^r  \prod_i y_i\left((\ell + r) - \sum_i m_i c\right)$ - we can assume this is nonzero since $c$ is generic and none of the $y_i$ are $0$. Therefore we can write $[z^\ell]a(z)$ as a linear combination of $[z^{\ell+1}]a(z),\dots,[z^{\ell+r}]a(z)$ when $\ell \le p-r$.

We see that since $[z^{p-r+1}]a(z)=\dots=[z^p]a(z)=0$, we can use induction starting with $\ell=p-r$ to show that $[z^\ell]a(z)=0$ for $\ell=p-r,p-r-1,\dots,0$. Therefore $[z^0]a(z)=0$; however, it is clear from the definition of $a(z)$ that this is not the case. Therefore we have a contradiction, so one of the $y_i$ is $0$.

We can assume without loss of generality that $y_{r-1}=0$. Then if $r>1$, we can go through the same argument again with $y_0,\dots,y_{r-2}$. We are able to do this because our calculation of $F'(z)$ and $V(z)$ so that $z^p \mid V(z)$ relies only on the fact that $\sum m_iy_i=0$, which will still be true if we leave out $y_{r-1}=0$. We will then find that one of the $y_i$ with $i < r-1$ is $0$, which gives a contradiction since the $y_i$ are distinct. Therefore we must have $r=1$ and $y_0=0$. 
\end{proof}

@@

\begin{theorem}\label{thm:main} The irreducible representation $A/J$  of $\HH_{\hbar,c}(\h)$ is a complete intersection with  Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$. The maximal ideal $J$ is generated by $[z^p]F_a(z)$ for the formal power series
\[
F_a(z)=\frac{1}{1-x_az} \prod_{j=0}^{n-1}\left( \sum_{k=0}^{p-1} \binom{c}{k}(-1+\prod_{j=0}^{n-1} (1-x_jz))^k\right) \text{ for $a=0,\dots,n-2$}.
\]
\end{theorem}
\begin{proof}
It suffices to show that the $f_a$ generate the ideal $J$ and that $A/J$ has Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$.

By Propositions~\ref{prop:linind} and \ref{prop:ci}, $A/I$ is a complete intersection with $n-1$ generators of degree $p$. It then must have Hilbert series $h_{A/I}(t)=\left(\frac{1-t^p}{1-t}\right)^{n-1}$. By Proposition~\ref{prop:ann}, the generators of $I$ are annihilated by the Dunkl operators, so $I \subseteq J$.

By \cite[Proposition 3.4]{BC1}, we see that the Hilbert series of $A/J$ is $\left(\frac{1-t^p}{1-t}\right)^{n-1}h(t^p)$ for some polynomial $h$ with nonnegative integer coefficients; since $I \subseteq J$, we see that $h_{A/I}(t) \ge h_{A/J}(t)$ coefficient-wise; however by this restriction of the form of $h_{A/J}(t)$, we see that the only possible choice for $h$ is $h(t)=1$. Therefore $h_{A/I}(t)=h_{A/J}(t)$, so $I=J$ and these $n-1$ generators generate the whole ideal $J$.
\end{proof}

>>>>

 \newtheorem*{thm:main}{Theorem \ref{thm:main}}
\begin{thm:main} The irreducible representation $A/J$  of $\HH_{\hbar,c}(\h)$ is a complete intersection with  Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$. The maximal ideal $J$ is generated by $[z^p]F_i(z)$ for the formal power series
\[
F_i(z)=\frac{1}{1-x_iz} \left( \sum_{k=0}^{p-1} \binom{c}{k}(-1+\prod_{j=0}^{n-1} (1-x_jz))^k\right) \text{ for $i=0,\dots,n-2$}.
\]
\end{thm:main}

@@

We first consider the case $p=2$. In this case $[z^2]F_j(z)=1-c$ and $[z^2]F_i(z)=-c$. By setting $x_j=1,x_{n-1}=-1$ and $x_i=0$ for $i \ne j, i \ne n-1$ for all $j=0,\dots,n-1$ in succession, we find that 
\[
\lambda_j-c\sum_{i=0}^{n-2} \lambda_i = 0 \text{ for all }j.
\]
Therefore for all $j$, $\lambda_j=c\sum_{i=0}^{n-2} \lambda_i$. Then in particular all of the $\lambda_i$ are equal to some $\lambda \in k$ with $(1-c(n-1))\lambda = 0$. Since $c$ is generic, we can assume $(1-c(n-1)) \ne 0$, so then $\lambda=0$. Therefore $\lambda_j=0$ for all $j$; then the $f_i$ for $i=0,\dots,n-2$ are linearly independent as desired.

We now suppose $p>2$. Then we have $[z^p]F_j(z)=f_j=\sum_{k=0}^{(p-1)/2} (-1)^k\binom{c}{k}=\binom{c-1}{(p-1)/2}$ and $[z^p]F_i(z)=f_i=0$ for $n-1>i \ne j$ (since $p > 2$). Since $c$ is generic, we can assume $\binom{c-1}{(p-1)/2} \ne 0$. Then since $\sum_{i=0}^{n-2} \lambda_if_i=0$, we see that $\lambda_j\binom{c-1}{(p-1)/2}=0$; then we must have $\lambda_j=0$.

@@

\begin{proposition}\label{prop:ci} Let $I \subseteq A$ be the ideal generated by $f_i$ for $i=0,\dots,n-2$.  Then $A/I$ is a complete intersection. 
\end{proposition}

@@

We suppose that the $x_i$ take the distinct values $\{y_0,\dots,y_{r-1}\}$ with multiplicity $m_i>0$. Then $g(z)=(1-y_iz)^{m_i}$. Recall that
\[
F(z)=\sum_{k=0}^{p-1} \binom{c}{k}(g-1)^k.
\]
Then we assume that $[z^p]\left(\frac{1}{1-y_iz}F(z)\right)=0$ for all $i=0,\dots,r-1$. We wish to show that $r=1$ and $y_0=0$.

We will assume that $\prod_i y_i \ne 0$ in order to derive a contradiction and show that one of the $y_i$ is $0$. Since $\sum_{i=0}^{n-1} x_i=0$ we also have $\sum_{i=0}^{r-1} m_iy_i=0$; this enables us to compute the derivative $F'(z)$, which is 
\[
F'(z)=V(z)+\left(\sum_j \frac{-cx_j}{1-x_jz}F(z)\right)=V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right),
\]
where $[z^\ell]V(z)=0$ for $0 \le \ell \le p$.

We will use a `residues by parts' argument; using the fact that the residue of an exact differential is zero, we will relate the coefficients of a power series to the coefficients of the derivative of that power series and derive a contradiction. 
Let
\[
a(z)=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z).
\]
We then see that
\begin{align*}
a'(z)&=\frac{d}{dz}\left(\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z)\right)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F'(z)+\sum_{j=0}^{r-1}\frac{y_j}{(1-y_jz)\prod_{i=0}^{r-1}(1-y_iz)}F(z)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}\left(V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}a(z)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}-\left(\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)\right).
\end{align*}
Since $[z^\ell]V(z)=0$ for $0 \le \ell \le p$ we see that $\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}=z^pb(z)$ for some power series $b(z)$. Therefore
\[
a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)
\]
for some power series $b(z)$.

@@

Note that for any $i$, $a(z)b_i(z)=\frac{1}{1-y_iz}F(z)$. Then in particular, we have 
\[
[z^p]a(z)b_i(z)=[z^p]\frac{1}{1-y_iz}F(z)=0
\]
Therefore for any $\lambda_i \in k$, we have
\begin{equation}\label{eq:lincomb}
\sum_{i=0}^{r-1} \lambda_i[z^p]a(z)b_i(z)=[z^p]\left( a(z)\sum_{i=0}^{r-1}\lambda_ib_i(z)\right)=0.
\end{equation}
Recall that the $b_i(z)$ are linearly independent and have degree at most $r-1$. Therefore for $0 \le k \le r-1$ we can choose $\{\lambda_i^k\}$ so that 
\[
\sum_i\lambda_i^kb_i(z)=z^k.
\]
Choosing these $\lambda_i^k$ in (\ref{eq:lincomb}), we see that $[z^p](a(z)z^k)=[z^{p-k}]a(z)=0$ for $k=0,\dots,r-1$.

We will now derive from this that $[z^0]a(z)=0$, a contradiction. If $r > p$ then this has already been established, so it suffices to assume $r \le p$.

For any integer $\ell$, define the Laurent polynomial 
\[
h_\ell(z)=z^{-\ell-r}\prod_i(1-y_iz).
\]
Notice that
\begin{align*}
\frac{d}{dz}( h_\ell(z) a(z)) &= -(l + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) + h_\ell(z) a'(z) \\
&=  -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_\ell(z) a(z) - h_\ell(z) \sum_j \frac{(m_j c - 1)y_i}{1 - y_ijz} a(z)  + z^ph_\ell(z) b(z)\\
&= -(\ell + r)z^{-1} h_\ell(z) a(z) - \sum_i \frac{y_im_i c}{1 - y_iz} h_\ell(z) a(z) + z^ph_\ell(z) b(z)\\
&= -\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z) a(z) + z^ph_\ell(z) b(z).
\end{align*}
We note that 
\[
-\left((\ell + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_\ell(z)
\]
is a Laurent polynomial with lowest degree term 
\[
- z^{-l-r-1} \left(\ell + r + \sum_i y_i m_i c\right) = - z^{-\ell-r-1}(\ell+ r)
\]
and highest degree term 
\[
- z^{-\ell - 1} (-1)^r  \prod_i y_i\left((\ell + r) - \sum_i m_i c\right).
\]
The lowest degree term of the formal power series $z^ph_\ell(z)b(z)$ has degree at least $p-\ell-r$. We note that we must have $[z^{-1}]\left(\frac{d}{dz}( h_\ell(z) a(z))\right)=0$, so for $p-\ell-r \ge 0$, we see that $[z^{-1}]\left(\frac{d}{dz}( h_\ell(z) a(z))\right)=0$ is a linear combination of $[z^\ell]a(z),[z^{\ell+1}]a(z),\dots,[z^{\ell+r}]a(z)$ in which $[z^\ell]a(z)$ has nonzero coefficient $ (-1)^r  \prod_i y_i\left((\ell + r) - \sum_i m_i c\right)$. The latter coefficient is nonzero since $c$ is generic and none of the $y_i$ are $0$. Therefore $[z^\ell]a(z)$ is a linear combination of $[z^{\ell+1}]a(z),\dots,[z^{\ell+r}]a(z)$ when $\ell \le p-r$.

Since $[z^{p-r+1}]a(z)=\dots=[z^p]a(z)=0$, by induction starting with $\ell=p-r$ we find that $[z^\ell]a(z)=0$ for $\ell=p-r,p-r-1,\dots,0$. Therefore, $[z^0]a(z)=0$; a contradiction. We conclude that $y_i=0$ for some $i$.

Assume without loss of generality that $y_{r-1}=0$. If $r>1$, we can repeat the same argument again with $y_0,\dots,y_{r-2}$.

We are able to do this because our calculation of $F'(z)$ and $V(z)$ so that $z^p \mid V(z)$ relies only on the fact that $\sum m_iy_i=0$. Since $y_{r-1}=0$, we see that $\sum_{i=0}^{r-2}m_iy_i=\sum_{i=0}^{r-1}m_iy_i=0$. Then since the $y_i$ are distinct, $\prod_{i=0}^{r-2} y_i \ne 0$, so both necessary assumptions are true. We will then find that one of the $y_i$ with $i < r-1$ is $0$, which gives a contradiction since the $y_i$ are distinct. Therefore we must have $r=1$ and $y_0=0$. 
\end{proof}

\subsection{Completing the proof}

@@

\begin{theorem}\label{thm:main} The irreducible representation $A/J$  of $\HH_{\hbar,c}(\h)$ is a complete intersection with  Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$. The maximal ideal $J$ is generated by $[z^p]F_i(z)$ for the formal power series
\[
F_i(z)=\frac{1}{1-x_iz} \left( \sum_{k=0}^{p-1} \binom{c}{k}(-1+\prod_{j=0}^{n-1} (1-x_jz))^k\right) \text{ for $i=0,\dots,n-2$}.
\]
\end{theorem}
\begin{proof}
It suffices to show that the $f_i$ generate the ideal $J$ and that $A/J$ has Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$.

By Propositions~\ref{prop:linind} and \ref{prop:ci}, $A/I$ is a complete intersection with $n-1$ linearly independent generators $f_i$ in degree $p$, so then its Hilbert series is $h_{A/I}(t)=\left(\frac{1-t^p}{1-t}\right)^{n-1}$. By Proposition~\ref{prop:ann}, the generators of $I$ are annihilated by the Dunkl operators, so $I \subseteq J$.

By \cite[Proposition 3.4]{BC1}, the Hilbert series of $A/J$ is $\left(\frac{1-t^p}{1-t}\right)^{n-1}h(t^p)$ for some polynomial $h$ with nonnegative integer coefficients. Since $I \subseteq J$, we see that $h_{A/I}(t) \ge h_{A/J}(t)$ coefficient-wise; hence $h(t)=1$. Therefore $h_{A/I}(t)=h_{A/J}(t)$, so $I=J$, completing the proof.
\end{proof}
